---
title: "Building Sustainability"
author: "Uine Kailamäki"
date: "2021"
output:
  html_document:
    theme: cosmo
    toc: true
    fig_caption: true
    fig_width: 12
    fig_height: 8
    code_folding: hide

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width=12, fig.height=8)
```

For the reader: You can click the black "code" box on the right hand side to either see or hide the R code behind analysis results and graphs. 

The results are arranged to separate tabs by site/area, so that it's easy to look at just the results of the site of interest, instead of scrolling accidentally down to a different site and getting confused.

The results for each site consist of four parts: pXRF, particle size distribution, loss on ignition, and colorimetry.

# Sites {.tabset}

## Ashdod-Yam 
```{r child = "results/ashdod-yam/index.Rmd"}
```
## Ashdod-Yam: Byzantine 
```{r child = "results/byzantine/index.Rmd"}
```
## Israel 
```{r child = "results/israel/index.Rmd"}
```
## Palaepaphos 
```{r child = "results/palaepaphos/index.Rmd"}
```
## Artaxata 
```{r child = "results/artaxata/index.Rmd"}
```

