---
author: "Uine Kailamäki"
output:
  html_document:
    theme: cosmo
    fig_caption: true
    fig_width: 12
    fig_height: 8 
    
---

From literature we can identify three rough classes of elements that could be used to analyze mudbricks from pXRF results: 

Good: Rb, Sr, Y, Zr, Nb, Th, Pb, Cu, Zn, Fe \
Neutral: V, Cr, Co, Ni, Ca, Ti \
Bad: Na, P, Ba (+ Ca, Al)

The "bad" elements were excluded from the pXRF results from the get-go. After scaling and averaging the remaining values we ended up with two analysis datasets: **"pxrf_all"** that contains all the elements that have non-zero results, and **"pxrf_no_na"**, in which all the elements that have even one NA result are omitted.  

```{r}
MB_table <- matrix(c("LA54/4",  "30 cm x 18 cm", "Top row - weathered",
                  "LA54/4",	"25 cm x 18 cm",	"Mid 4th row from the top",
                  "LA54/4",	"35 cm x 18 cm (est.)", "B. 8th row from the top",
                  "LA54/4",	"45 cm x 18 cm",	"9th row from the top",
                  "LA54/4",	"46 cm x 16 cm",	"6th row from the top",
                  "LA59/2",	"45 cm x 15 cm (est.)",	"Top row - weathered",
                  "LA59/2",	"45 cm x 15 cm (est.)",	"3rd row from the top",
                  "LA59/2",	"33 cm x 15 cm (est.)",	"5th row below stones",
                  "LA59/2",	"44 cm x 13 cm",	"9th row of the top",
                  "LA59/2",	"40 cm x 20 cm",	"B. below disturbance under the marl",
                  "LA54/7",	"14 cm x 15 cm (est.)", "(no additional info)"), 
                  ncol=3, byrow=TRUE)
colnames(MB_table) <- c("Context", "Size", "Row")
rownames(MB_table) <- c("PP-9", "PP-10", "PP-11", "PP-12", "PP-13", "PP-14", "PP-15", "PP-16", "PP-17", "PP-18", "PP-19")
MB_table <- as.table(MB_table)

library(knitr)
kable(MB_table)
```

#### Data preparation

```{r}
# Libraries
library(openxlsx); 
library(dplyr);
library(ggfortify); # Pretty PCA plots
library(FactoMineR); # Fast automatic PCA graphs
library(factoextra) # Eigenvalues for PCA
library(dendextend); # Dendrograms
library(plot3D); # 3D plot for PCA results

# Read the raw data
pxrf <- read.xlsx("data/pxrf_PP.xlsx", sep=";")

# Remove any rows containing words 'ERROR' or 'TEST'
pxrf <- filter(pxrf, !grepl('TEST', Sample))
pxrf <- filter(pxrf, !grepl('ERROR', Sample))

# Remove unused columns
pxrf <- pxrf %>% 
        select(-Site) %>% 
        select(-'Other;name') %>% 
        select(-'File;#') %>% 
        select(-DateTime) %>% 
        select(-Application) %>% 
        select(-Method) %>% 
        select(-ElapsedTime) %>% 
        select(-'Cal;Check')
  
# Coercing "< LOD" -results to NA:s  
pxrf[-c(1:3)] <- pxrf[-c(1:3)] %>% mutate_if(is.character,as.numeric)

# Transforming NA values to zeros
pxrf[is.na(pxrf)] <- 0

# Averaging results by "Sample", "Area" and "Type" columns
pxrf_averaged <- aggregate(pxrf, by = list(pxrf$Sample, pxrf$Area, pxrf$Type), FUN = mean)

# Removing old (now empty) columns
pxrf_averaged <- pxrf_averaged %>% select(-Sample)
pxrf_averaged <- pxrf_averaged %>% select(-Area)
pxrf_averaged <- pxrf_averaged %>% select(-Type)

# Assigning sample names as new row names 
rownames(pxrf_averaged) <- pxrf_averaged$'Group.1'
pxrf_averaged <- pxrf_averaged %>% select(-Group.1)

# Renaming the other two newly created groups back to "Type" and "Area"
names(pxrf_averaged)[names(pxrf_averaged) == "Group.2"] <- "Area"
names(pxrf_averaged)[names(pxrf_averaged) == "Group.3"] <- "Type"

# Removing error columns from the scaled analysis data set
pxrf_scaled <- pxrf_averaged %>% select(-ends_with("Err"))

# Scaling the data with standard z-score method
pxrf_scaled[-c(1:2)] <- scale(pxrf_scaled[-c(1:2)])

# Removing columns that only contain NA values
all_na <- function(x) any(!is.na(x))
pxrf_all <- pxrf_scaled %>% select_if(all_na)

```

**pxrf_all**: All possible elements
**pxrf_errors**: Means of  available values and error values for a quick glimpse

```{r}
colnames(is.na(pxrf_all[-c(1:2)]))

# Mean values and mean errors
pxrf_errors <- pxrf_averaged %>% select_if(all_na)
pxrf_errors <- colMeans(pxrf_errors[-c(1:2)])
pxrf_errors <- (pxrf_errors * 1000)
pxrf_errors <- round(pxrf_errors, digits = 2)

# Combining the individual measurement errors properly
pxrf_errors2 <- pxrf[,grepl("Err$", names(pxrf))]

### 
### errorfunc <- function(x){(1/length(x))*(sqrt(sum(x^2)))}
errorfunc <- function(x){(sqrt(sum(x^2)))}
pxrf_errors2<- aggregate(pxrf_errors2, by = list(pxrf$Sample), FUN = errorfunc)

rownames(pxrf_errors2) <- pxrf_errors2$'Group.1'
pxrf_errors2 <- pxrf_errors2 %>% select(-Group.1)

# Dataframe with correct combined errors for each sample
pxrf_errors2

# Replacing mean errors with calculated errors
pxrf_averaged[names(pxrf_errors2)] <- pxrf_errors2

# Printing table with errors for selected elements
pxrf_table <- pxrf_averaged[c(19:22,25:28,35:36,41:48)]
pxrf_table <-  (pxrf_table * 1000)
pxrf_table <- round(pxrf_table, digits = 2)
kable(pxrf_table)
```

**pxrf_final**: Final analysis data set with only the selected elements

```{r}
# Dropping the dubious elements and elements with too many NA values
pxrf_final <- pxrf_all %>% 
        select(-Al2O3) %>% 
        select(-CaO) %>% 
        select(-SiO2) %>%
        select(-Cl) %>%  
        select(-K2O) %>% 
        select(-Ni) %>% 
        select(-P2O5) %>%  
        select(-S) %>% 
        select(-As) %>% 
        select(-Rh) %>% 
        select(-Ag) %>% 
        select(-MgO) %>% 
        select(-Cu) %>% 
        select(-Cr) %>% 
        select(-Nb) %>% 
        select(-Mo) %>% 
        select(-Ba) %>% 
        select(-Ta) %>% 
        select(-Pb)

# Transforming remaining NA values to zeros for analysis
pxrf_final[is.na(pxrf_final)] <- 0

# Final analysis data set        
colnames(pxrf_final[-c(1:2)])

```

**pxrf_MB**: Final analysis data set for MUDBRICK samples only (= no soil)

```{r}
pxrf_MB <- pxrf_final[c(1:11), ]
```

#### pXRF: K means cluster analysis

Only mudbrick samples, K-means cluster analysis with 5 possible clusters

```{r}
# PC autoplot with K-means clusters
set.seed(1)
autoplot(kmeans(pxrf_MB[3:10], 4), data = pxrf_MB, label = TRUE, label.size = 3)

```

#### pXRF: PCA

**PCA with only mudbrick samples**: 

```{r}
# PCA
PCA_01 <- PCA(pxrf_MB[3:10], graph = FALSE)

# Eigenvalues
get_eigenvalue(PCA_01)
fviz_eig(PCA_01)

# Contributions of variables to PC1 and PC2
fviz_contrib(PCA_01, choice = "var", axes = 1, top = 10)
fviz_contrib(PCA_01, choice = "var", axes = 2, top = 10)

# Graph of variables with contribution indication
fviz_pca_var(PCA_01, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE # Avoid text overlapping
             )

# Graph of individuals with contribution indication
fviz_pca_ind(PCA_01,
             col.ind = pxrf_MB$Area, # color by groups
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             legend.title = "Context")

#3D plot
scores = as.data.frame(PCA_01$ind)
scatter3D(scores$'coord.Dim.2', scores$'coord.Dim.3', scores$'coord.Dim.1', 
          xlab = "PC2 (16,29 %)", ylab = "PC3 (5,95 %)", zlab="PC1 (71,83 %)",
          phi = 0, bty = "g",  type = "h", 
          ticktype = "detailed", pch = 19, cex = 1.2)
text3D(scores$'coord.Dim.2', scores$'coord.Dim.3', scores$'coord.Dim.1', labels = rownames(scores),
          add = TRUE, colkey = FALSE, cex = 0.7)


```

**PCA with soil samples**: \

```{r}
# PCA
PCA_02 <- PCA(pxrf_final[3:10], graph = FALSE)

# Eigenvalues
get_eigenvalue(PCA_02)
fviz_eig(PCA_02)

# Contributions of variables to PC1 and PC2
fviz_contrib(PCA_02, choice = "var", axes = 1, top = 10)
fviz_contrib(PCA_02, choice = "var", axes = 2, top = 10)
fviz_contrib(PCA_02, choice = "var", axes = 3, top = 10)

# Graph of variables with contribution indication
fviz_pca_var(PCA_02, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE # Avoid text overlapping
             )

# Graph of individuals
fviz_pca_ind(PCA_02,
             col.ind = pxrf_final$Area,
             legend.title = "Context")

# Graph of individuals
fviz_pca_ind(PCA_02,
             col.ind = pxrf_final$Type, # color by groups
             legend.title = "Type")

#3D plot
scores = as.data.frame(PCA_02$ind)
scatter3D(scores$'coord.Dim.2', scores$'coord.Dim.3', scores$'coord.Dim.1', 
          xlab = "PC2 (19,44 %)", ylab = "PC3 (16,42 %)", zlab="PC1 (55,68 %)",
          phi = 0, bty = "g",  type = "h", 
          ticktype = "detailed", pch = 19, cex = 1.2)
text3D(scores$'coord.Dim.2', scores$'coord.Dim.3', scores$'coord.Dim.1', labels = rownames(scores),
          add = TRUE, colkey = FALSE, cex = 0.7)

```

#### pXRF: HCA

**HCA including only MB samples, colored by area**: \

```{r}
# HCA dendrogram, samples color coded by area:
dend <- 
    pxrf_MB %>%                         # data
    dist %>%                            # calculate a distance matrix
    hclust(method = "ward.D2") %>%      # hierarchical clustering 
    as.dendrogram %>%                   # turn the object into a dendrogram
    highlight_branches_col

    area <- as.factor(pxrf_MB[, 1])
    n_area <- length(unique(area))
    cols_a <- colorspace::rainbow_hcl(n_area, c = 70, l  = 50)
    col_area <- cols_a[area]
    labels_colors(dend) <- col_area[order.dendrogram(dend)]

    plot(dend, main="HCA with sample areas")
    rect.dendrogram(dend, 6, border = "Grey", lty = 5)
    legend("topright", legend = levels(area), fill = cols_a)
    

# PCA biplot with HCA groups
    
# Compute PCA with ncp = 3
res.pca <- PCA(pxrf_MB[3:10], ncp = 3, graph = FALSE)
# Compute hierarchical clustering on principal components
res.hcpc <- HCPC(res.pca, graph = FALSE)
    
fviz_dend(res.hcpc, 
          cex = 0.7,                     # Label size
          palette = "jco",               # Color palette see ?ggpubr::ggpar
          rect = TRUE, rect_fill = TRUE, # Add rectangle around groups
          rect_border = "jco",           # Rectangle color
          labels_track_height = 0.8      # Augment the room for labels
          )
      
  fviz_cluster(res.hcpc,
             repel = TRUE,            # Avoid label overlapping
             show.clust.cent = TRUE, # Show cluster centers
             palette = "jco",         # Color palette see ?ggpubr::ggpar
             ggtheme = theme_minimal(),
             main = "Factor map"
             )

```

**HCA with soil samples, colored by type**: \

```{r}
# HCA dendrogram, samples color coded by type:
dend3 <- 
    pxrf_final %>%                      # data
    dist %>%                            # calculate a distance matrix
    hclust(method = "ward.D2") %>%      # hierarchical clustering 
    as.dendrogram %>%                   # turn the object into a dendrogram
    highlight_branches_col

    type <- as.factor(pxrf_final[, 2])
    n_type <- length(unique(type))
    cols_t <- colorspace::rainbow_hcl(n_type, c = 70, l  = 50)
    col_type <- cols_t[type]
    labels_colors(dend3) <- col_type[order.dendrogram(dend3)]

    plot(dend3, main="HCA with all samples by type")
    rect.dendrogram(dend3, 2, border = "Grey", lty = 5)
    legend("topright", legend = levels(type), fill = cols_t)
    
    
    # PCA biplot with HCA groups
    
# Compute PCA with ncp = 3
res.pca <- PCA(pxrf_final[3:10], ncp = 5, graph = FALSE)
# Compute hierarchical clustering on principal components
res.hcpc <- HCPC(res.pca, graph = FALSE)
    
fviz_dend(res.hcpc, 
          cex = 0.7,                     # Label size
          palette = "jco",               # Color palette see ?ggpubr::ggpar
          rect = TRUE, rect_fill = TRUE, # Add rectangle around groups
          rect_border = "jco",           # Rectangle color
          labels_track_height = 0.8      # Augment the room for labels
          )

```


