---
title: "Palaepaphos mudbrick analysis"
author: "Uine Kailamäki"
output:
  html_document:
    theme: cosmo
    fig_caption: true
    fig_width: 12
    fig_height: 8 
    
---


Introduction

```{r}
# Libraries
library(dplyr); 
library(GGally); 
library(corrplot); 
library(tidyr); 
library(openxlsx); 
library(cluster); 
library(ggplot2); 
library(ggfortify); 
library(pca3d);
library(FactoMineR); # fast PCA graphs
library(ggdendro); # dendrograms together with ggplot
library(dendextend); # dendrograms - change parameters
library(tidyverse) # rectangles around HClust

```

Data preparation

```{r}
# Read data
pxrf <- read.xlsx("data/pxrf_PP.xlsx", sep=";")

# Remove any rows containing words 'ERROR' or 'TEST'
pxrf <- filter(pxrf, !grepl('TEST', Sample))
pxrf <- filter(pxrf, !grepl('ERROR', Sample))

# Turn any "LOD":s and missing values to "NA":s to allow scaling
values <- c("MgO", "Al2O3", "SiO2", 
            "S", "Cl", "K2O", "CaO", "Ti", "V", "Cr", 
            "Mn", "Fe", "Co", "Ni", "Cu", "Zn", "As", 
            "Se", "Rb", "Sr", "Y", "Zr", "Nb", "Mo", 
            "Rh", "Pd", "Ag", "Cd", "Sn", "Sb", 
            "La", "Ce", "Hf", "Ta", "W", "Pt", "Au", 
            "Hg", "Tl", "Pb", "Bi", "Th", "U")

pxrf_NA <- select(pxrf, one_of(values))
pxrf_NA <- pxrf_NA %>% mutate_if(is.character,as.numeric)

# Average data by "Sample", "Area" and "Type" columns from the original dataset
average_pxrf <- aggregate(pxrf_NA, by = list(pxrf$Sample, pxrf$Area, pxrf$Type), FUN = mean)

# Assign sample names as row names, and rename the other newly created columns back to "Type" and "Area" for clarity
rownames(average_pxrf) <- average_pxrf$Group.1
average_pxrf <- average_pxrf %>% select(-Group.1)
names(average_pxrf)[names(average_pxrf) == "Group.2"] <- "Area"
names(average_pxrf)[names(average_pxrf) == "Group.3"] <- "Type"

# Change "Type" and "Area" to factors
average_pxrf$Type <- factor(average_pxrf$Type)
average_pxrf$Area <- factor(average_pxrf$Area)

# Removing columns that only have NA values
all_na <- function(x) any(!is.na(x))
pxrf_all <- average_pxrf %>% select_if(all_na)

# Removing columns if they contain any NA values at all
any_na <- function(x) all(!is.na(x))
pxrf_no_na <- average_pxrf %>% select_if(any_na)

# Datasets with all the samples, including soil and old mudbrick samples
summary(pxrf_no_na)
summary(pxrf_all)

# Datasets with only our new mudbrick samples (no soil or previously published ones)
MB_nona <- pxrf_no_na[c(1:11), ]
MB_all <- pxrf_all[c(1:11), ]

# Removing As, Nb, Rh, Ag and Pb as they have almost no viable measurement values
MB_all <- MB_all %>% 
        select(-As) %>% 
        select(-Nb) %>% 
        select(-Rh) %>%   
        select(-Ag) %>% 
        select(-Pb) 

summary(MB_nona)
summary(MB_all)

```

Elemental correlations in new mudbrick samples

```{r}
# Visualize correlations between elements
cor(MB_all[3:20]) %>% corrplot (type = "lower", tl.cex = 1, tl.pos="d", cl.pos="r")
```

K-means cluster analysis

```{r}

# Optimal number of clusters in K-means analysis: 2
k_max <- 8
twcss <- sapply(1:k_max, function(k){kmeans(MB_nona[3:8], k)$tot.withinss})
qplot(x = 1:k_max, y = twcss, geom = 'line')

# K-means clustering in ggpairs: 2 clusters
km <-kmeans(MB_nona[3:8], centers = 2)
cluster_x <- as.factor(km$cluster)
ggpairs(MB_nona[3:8], mapping = aes(col = cluster_x))

```

### pXRF: PCA

PCA with only the no-NA:s elements

```{r}

# PCA for only non-NA elements
colnames(MB_nona)
pca_1 <- prcomp(MB_nona[3:8], scale = TRUE, center = TRUE)
summary(pca_1)

biplot(pca_1, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"))
autoplot(pca_1, data=MB_nona, colour="red", shape = FALSE, label = TRUE,  main = "PCA Palaepaphos 1: no NA:s at all")
PCA(MB_nona[3:8])

```

PCA with all the feasible elements

```{r}
# NA:s converted to zeros
MB_all[is.na(MB_all)] <- 0

colnames(MB_all)
pca_2 <- prcomp((MB_all[3:20]), scale = TRUE, center = TRUE)
summary(pca_2)

biplot(pca_2, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"))
autoplot(pca_2, data=MB_all, colour="red", shape = FALSE, label = TRUE,  main = "PCA Palaepaphos 2: NA:s permitted")
PCA(MB_all[3:20])

```

### PCA with all the Mudbrick and soil samples (including the previously published ones): No NA values permitted

```{r}
# PCA for only non-NA elements
colnames(pxrf_no_na)
pca_3 <- prcomp(na.omit(pxrf_no_na[3:6]), scale = TRUE, center = TRUE)
summary(pca_3)

biplot(pca_3, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"))
autoplot(pca_3, data=pxrf_no_na, colour='Area', shape = FALSE, label = TRUE,  main = "PCA Palaepaphos grouped by area")
autoplot(pca_3, data=pxrf_no_na, colour='Type', shape = FALSE, label = TRUE,  main = "PCA Palaepaphos grouped by sample type")

```

### PCA with all the Mudbrick samples 

(including the previously published ones): NA values are permitted
This is not feasible as the difference in what values are available is too different between old and new samples.

```{r}
# NA:s converted to zeros
pxrf_all[is.na(pxrf_all)] <- 0

colnames(pxrf_all)
pca_4 <- prcomp(pxrf_all[3:25], scale = TRUE, center = TRUE)
summary(pca_4)

biplot(pca_4, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"))
autoplot(pca_4, data=pxrf_all, colour='Area', shape = FALSE, label = TRUE,  main = "PCA Palaepaphos grouped by area")
autoplot(pca_4, data=pxrf_all, colour='Type', shape = FALSE, label = TRUE,  main = "PCA Palaepaphos grouped by sample type")

```

### HCA analysis

HCA including only new mudbrick samples: 

```{r}
# HCA dendrogam 1
dend <- 
    MB_all[3:20] %>%           # data
    dist %>%                            # calculate a distance matrix
    hclust(method = "ward.D2") %>%      # hierarchical clustering 
    as.dendrogram %>%                   # turn the object into a dendrogram
    set("branches_k_color", k=5) %>% 
    set("branches_lwd", 1.2) %>%
    set("labels_col", c("red", "black", "blue"), k=5) %>%    
    set("labels_cex", 1) %>% 
    set("leaves_pch", NA) 
# plot the dend in usual "base" plotting engine:
plot(dend)

# HCA dendrogam 2
HCA.ward5 <- hclust(dist(MB_all[3:20], method="euclid"), method="ward.D2")
plot(HCA.ward5, main="Ward's method")
rect.hclust(HCA.ward5, k=5)

# Divisive hierarchical clustering (DIANA) just for comparison
hc4 <- diana(MB_all[3:20])
pltree(hc4, cex = 1, hang = -1, main = "Dendrogram of diana")
# Divise coefficient; amount of clustering structure found
hc4$dc

```

### HCA analysis

HCA including all mudbrick samples, only no NA:s elements: 

```{r}
# HCA dendrogam 1
dend <- 
    pxrf_no_na[1:45, 3:8] %>%           # data
    dist %>%                            # calculate a distance matrix
    hclust(method = "ward.D2") %>%      # hierarchical clustering 
    as.dendrogram %>%                   # turn the object into a dendrogram
    set("branches_k_color", k=5) %>% 
    set("branches_lwd", 1.2) %>%
    set("labels_col", c("red", "black", "blue"), k=5) %>%    
    set("labels_cex", 1) %>% 
    set("leaves_pch", NA) 
# plot the dend in usual "base" plotting engine:
plot(dend)

# HCA dendrogam 2
HCA.ward5 <- hclust(dist(pxrf_no_na[1:45, 3:8], method="euclid"), method="ward.D2")
plot(HCA.ward5, main="Ward's method")
rect.hclust(HCA.ward5, k=5)

# Divisive hierarchical clustering (DIANA) just for comparison
hc4 <- diana(pxrf_no_na[3:8])
pltree(hc4, cex = 1, hang = -1, main = "Dendrogram of diana")
# Divise coefficient; amount of clustering structure found
hc4$dc

```


