---
author: "Uine Kailamäki"
output:
  html_document:
    theme: cosmo
    fig_caption: true
    fig_width: 12
    fig_height: 8 
    
---

From literature we can identify three rough classes of elements that could be used to analyze mudbricks from pXRF results: 

Good: Rb, Sr, Y, Zr, Nb, Th, Pb, Cu, Zn, Fe \
Neutral: V, Cr, Co, Ni, Ca, Ti \
Bad: Na, P, Ba (+ Ca, Al)

The "bad" elements were excluded from the pXRF results from the get-go. After scaling and averaging the remaining values we ended up with two analysis datasets: **"pxrf_all"** that contains all the elements that have non-zero results, and **"pxrf_no_na"**, in which all the elements that have even one NA result are omitted.  

```{r}
MB_table <- matrix(c("LA54/4",  "30 cm x 18 cm", "Top row - weathered",
                  "LA54/4",	"25 cm x 18 cm",	"Mid 4th row from the top",
                  "LA54/4",	"35 cm x 18 cm (est.)", "B. 8th row from the top",
                  "LA54/4",	"45 cm x 18 cm",	"9th row from the top",
                  "LA54/4",	"46 cm x 16 cm",	"6th row from the top",
                  "LA59/2",	"45 cm x 15 cm (est.)",	"Top row - weathered",
                  "LA59/2",	"45 cm x 15 cm (est.)",	"3rd row from the top",
                  "LA59/2",	"33 cm x 15 cm (est.)",	"5th row below stones",
                  "LA59/2",	"44 cm x 13 cm",	"9th row of the top",
                  "LA59/2",	"40 cm x 20 cm",	"B. below disturbance under the marl",
                  "LA54/7",	"14 cm x 15 cm (est.)", "(no additional info)"), 
                  ncol=3, byrow=TRUE)
colnames(MB_table) <- c("Context", "Size", "Row")
rownames(MB_table) <- c("PP-9", "PP-10", "PP-11", "PP-12", "PP-13", "PP-14", "PP-15", "PP-16", "PP-17", "PP-18", "PP-19")
MB_table <- as.table(MB_table)

library(knitr)
kable(MB_table)
```

#### Data preparation

```{r}
# Libraries
library(openxlsx); 
library(dplyr);
library(ggfortify); # PCA autoplots
library(FactoMineR); # Fast PCA graphs
library(dendextend); # Dendrograms

# Read the raw data
pxrf <- read.xlsx("data/pxrf_PP.xlsx", sep=";")

# Remove any rows containing words 'ERROR' or 'TEST'
pxrf <- filter(pxrf, !grepl('TEST', Sample))
pxrf <- filter(pxrf, !grepl('ERROR', Sample))

# Remove unused columns
pxrf <- pxrf %>% 
        select(-Site) %>% 
        select(-'Other;name') %>% 
        select(-'File;#') %>% 
        select(-DateTime) %>% 
        select(-Application) %>% 
        select(-Method) %>% 
        select(-ElapsedTime) %>% 
        select(-'Cal;Check')
  
# Coercing "< LOD" -results to NA:s  
pxrf[-c(1:3)] <- pxrf[-c(1:3)] %>% mutate_if(is.character,as.numeric)

# Transforming NA values to zeros
pxrf[is.na(pxrf)] <- 0

# Averaging results by "Sample", "Area" and "Type" columns
pxrf_averaged <- aggregate(pxrf, by = list(pxrf$Sample, pxrf$Area, pxrf$Type), FUN = mean)

# Removing old (now empty) columns
pxrf_averaged <- pxrf_averaged %>% select(-Sample)
pxrf_averaged <- pxrf_averaged %>% select(-Area)
pxrf_averaged <- pxrf_averaged %>% select(-Type)

# Assigning sample names as new row names 
rownames(pxrf_averaged) <- pxrf_averaged$'Group.1'
pxrf_averaged <- pxrf_averaged %>% select(-Group.1)

# Renaming the other two newly created groups back to "Type" and "Area"
names(pxrf_averaged)[names(pxrf_averaged) == "Group.2"] <- "Area"
names(pxrf_averaged)[names(pxrf_averaged) == "Group.3"] <- "Type"

# Removing error columns from the scaled analysis data set
pxrf_scaled <- pxrf_averaged %>% select(-ends_with("Err"))

# Scaling the data with standard z-score method
pxrf_scaled[-c(1:2)] <- scale(pxrf_scaled[-c(1:2)])

# Removing columns that only contain NA values
all_na <- function(x) any(!is.na(x))
pxrf_all <- pxrf_scaled %>% select_if(all_na)

```

**pxrf_averaged**: All averaged data including averaged errors (Uine's note: for later use)
**pxrf_errors**: Averaged data including averaged errors, missing and dubious elements emitted

```{r}
pxrf_errors <- pxrf_averaged %>% select_if(all_na)
pxrf_errors <- pxrf_errors %>% 
        select(-Al2O3) %>% 
        select(-'Al2O3;Err') %>%   
        select(-CaO) %>% 
        select(-'CaO;Err') %>% 
        select(-SiO2) %>%
        select(-'SiO2;Err') %>%
        select(-Cl) %>%  
        select(-'Cl;Err') %>% 
        select(-K2O) %>% 
        select(-'K2O;Err') %>% 
        select(-Ni) %>% 
        select(-'Ni;Err') %>%   
        select(-P2O5) %>%  
        select(-'P2O5;Err') %>%   
        select(-S) %>% 
        select(-'S;Err') %>%  
        select(-V) %>% 
        select(-'V;Err') %>%   
        select(-As) %>% 
        select(-'As;Err') %>%   
        select(-Rh) %>% 
        select(-'Rh;Err') %>%   
        select(-Ag) %>%   
          select(-'Cr;Err') %>% 
          select(-'Co;Err') %>% 
          select(-'Se;Err') %>% 
          select(-Nb) %>% 
          select(-'Nb;Err') %>% 
          select(-'Mo;Err') %>% 
          select(-'Pd;Err') %>% 
          select(-'Cd;Err') %>% 
          select(-'Sn;Err') %>% 
          select(-'Sb;Err') %>% 
          select(-'Ba;Err') %>% 
          select(-'La;Err') %>% 
          select(-'Ag;Err') %>% 
          select(-'Ce;Err') %>% 
          select(-'Hf;Err') %>% 
          select(-'Ta;Err') %>% 
          select(-'W;Err') %>% 
          select(-'Pt;Err') %>% 
          select(-'Au;Err') %>% 
          select(-'Hg;Err') %>% 
          select(-'Tl;Err') %>% 
          select(-Pb) %>% 
          select(-'Pb;Err') %>% 
          select(-'Bi;Err') %>% 
          select(-'Th;Err') %>% 
          select(-'U;Err')

pxrf_errors[3:22] <- (pxrf_errors[3:22] * 1000)
pxrf_errors[3:22] <- round(pxrf_errors[3:22], digits = 1)

kable(pxrf_errors[3:22])

```

**pxrf_all**: All "possible" elements (the numbers below are the amount of NA values)

```{r}
colSums(is.na(pxrf_all[-c(1:2)]))
```

**pxrf_final**: Final analysis data set with only the selected elements

```{r}
# Dropping the dubious elements and elements with too many NA values
pxrf_final <- pxrf_all %>% 
        select(-Al2O3) %>% 
        select(-CaO) %>% 
        select(-SiO2) %>%
        select(-Cl) %>%  
        select(-K2O) %>% 
        select(-Ni) %>% 
        select(-P2O5) %>%  
        select(-S) %>% 
        select(-As) %>% 
        select(-Rh) %>% 
        select(-Ag) %>% 
        select(-MgO) %>% 
        select(-Cu) %>% 
        select(-Cr) %>% 
        select(-Nb) %>% 
        select(-Mo) %>% 
        select(-Ba) %>% 
        select(-Ta) %>% 
        select(-Pb)

# Transforming remaining NA values to zeros for analysis
pxrf_final[is.na(pxrf_final)] <- 0

# Final analysis data set        
colnames(pxrf_final[-c(1:2)])
```

**pxrf_MB**: Final analysis data set for MUDBRICK samples only (= no soil)

```{r}
pxrf_MB <- pxrf_final[c(1:11), ]
```

#### pXRF: K means cluster analysis

Only mudbrick samples, K-means cluster analysis with 5 possible clusters

```{r}
# PC autoplot with K-means clusters
set.seed(1)
autoplot(kmeans(pxrf_MB[3:10], 5), data = pxrf_MB, label = TRUE, label.size = 3)

```

#### pXRF: PCA

**PCA with only mudbrick samples**: 

```{r}
# Elements
colnames(pxrf_MB[-c(1:2)])

# PCA analysis
pca_1 <- prcomp(pxrf_MB[3:10])
summary(pca_1)

# PCA plots
biplot(pca_1, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"), main = "PCA Palaepaphos elements")

autoplot(pca_1, data=pxrf_MB, colour='Area', shape = FALSE, label = TRUE,  main = "PCA Palaepaphos grouped by area")

PCA(pxrf_MB[3:10])

#3D plot
scores = as.data.frame(pca_1$x)
scatter3D(scores$PC2, scores$PC3, scores$PC1, 
          xlab = "PC2 (23,6 %)", ylab = "PC3 (5,5 %)", zlab="PC1 (64,2 %)",
          phi = 0, bty = "g",  type = "h", 
          ticktype = "detailed", pch = 19, cex = 1.2)
text3D(scores$PC2, scores$PC3, scores$PC1, labels = rownames(scores),
          add = TRUE, colkey = FALSE, cex = 0.7)


```

**PCA with soil samples**: \

```{r}
# Elements
colnames(pxrf_final[3:10])

# PCA analysis
pca_3 <- prcomp(pxrf_final[3:10])
summary(pca_3)

# PCA plots
biplot(pca_3, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"), main = "PCA Palaepaphos elements")

autoplot(pca_3, data=pxrf_final, colour='Area', shape = FALSE, label = TRUE,  main = "PCA Palaepaphos grouped by area")

autoplot(pca_3, data=pxrf_final, colour='Type', shape = FALSE, label = TRUE,  main = "PCA Palaepaphos grouped by sample type")

PCA(pxrf_final[3:10])

```

#### pXRF: HCA

**HCA including only MB samples, colored by area**: \

```{r}
# HCA dendrogram, samples color coded by area:
dend <- 
    pxrf_MB %>%                         # data
    dist %>%                            # calculate a distance matrix
    hclust(method = "ward.D2") %>%      # hierarchical clustering 
    as.dendrogram %>%                   # turn the object into a dendrogram
    highlight_branches_col

    area <- as.factor(pxrf_MB[, 1])
    n_area <- length(unique(area))
    cols_a <- colorspace::rainbow_hcl(n_area, c = 70, l  = 50)
    col_area <- cols_a[area]
    labels_colors(dend) <- col_area[order.dendrogram(dend)]

    plot(dend, main="HCA with sample areas")
    rect.dendrogram(dend, 5, border = "Black", lty = 5)
    legend("topright", legend = levels(area), fill = cols_a)

```

**HCA with soil samples, colored by type**: \

```{r}
# HCA dendrogram, samples color coded by type:
dend3 <- 
    pxrf_final %>%                      # data
    dist %>%                            # calculate a distance matrix
    hclust(method = "ward.D2") %>%      # hierarchical clustering 
    as.dendrogram %>%                   # turn the object into a dendrogram
    highlight_branches_col

    type <- as.factor(pxrf_final[, 2])
    n_type <- length(unique(type))
    cols_t <- colorspace::rainbow_hcl(n_type, c = 70, l  = 50)
    col_type <- cols_t[type]
    labels_colors(dend3) <- col_type[order.dendrogram(dend3)]

    plot(dend3, main="HCA with all samples by type")
    rect.dendrogram(dend3, 7, border = "Black", lty = 5)
    legend("topright", legend = levels(type), fill = cols_t)

```


