---
title: "Ashdod-Yam mudbrick analysis"
author: "Uine Kailamäki"
output:
  html_document:
    theme: cosmo
    fig_caption: true
    fig_width: 12
    fig_height: 8 
    code_folding: hide
  
---


### pXRF

From literature we can identify three rough classes of elements that could be used to analyze mudbricks from pXRF results: 

Good: Rb, Sr, Y, Zr, Nb, Th, Pb, Cu, Zn, Fe \
Neutral: V, Cr, Co, Ni, Ca, Ti \
Bad: Na, P, Ba (+ Ca, Al)

The "bad" elements were excluded from the pXRF results from the get-go. After scaling and averaging the remaining values we ended up with two analysis datasets: **"pxrf_all"** that contains all the elements that have non-zero results, and **"pxrf_no_na"**, in which all the elements that have even one NA result are omitted.  

We end up with 20 elements in the "pxrf_all" dataset, and 12 elements in the "pxrf_no_na" dataset.


```{r}
# Libraries
library(dplyr); 
library(GGally); 
library(corrplot); 
library(tidyr); 
library(openxlsx); 
library(cluster); 
library(ggplot2); 
library(ggrepel);
library(ggfortify); 
library(pca3d);
library(ggdendro); # dendrograms together with ggplot
library(dendextend); # dendrograms - change parameters
library(ggtern) # ternary plots

# Read the raw data
pxrf <- read.xlsx("data/pxrf_AY.xlsx", sep=";")

# Remove any rows containing words 'ERROR' or 'TEST'
pxrf <- filter(pxrf, !grepl('TEST', Sample))
pxrf <- filter(pxrf, !grepl('ERROR', Sample))

# Turn any "LOD":s and missing values to "NA":s to allow scaling
values <- c("MgO", "Al2O3", "SiO2", 
            "S", "Cl", "K2O", "CaO", "Ti", "V", "Cr", 
            "Mn", "Fe", "Co", "Ni", "Cu", "Zn", "As", 
            "Se", "Rb", "Sr", "Y", "Zr", "Nb", "Mo", 
            "Rh", "Pd", "Ag", "Cd", "Sn", "Sb", 
            "La", "Ce", "Hf", "Ta", "W", "Pt", "Au", 
            "Hg", "Tl", "Pb", "Bi", "Th", "U")

pxrf_NA <- select(pxrf, one_of(values))
pxrf_NA <- pxrf_NA %>% mutate_if(is.character,as.numeric)

# Scaling the data with standard z-score method
scaled_pxrf <- as.data.frame(scale(pxrf_NA))

# Average data by "Sample", "Area" and "Type" columns from the original dataset
average_pxrf <- aggregate(scaled_pxrf, by = list(pxrf$Sample, pxrf$Area, pxrf$Type), FUN = mean)

# Assign sample names as row names, and rename the other newly created columns back to "Type" and "Area" for clarity
rownames(average_pxrf) <- average_pxrf$Group.1
average_pxrf <- average_pxrf %>% select(-Group.1)
names(average_pxrf)[names(average_pxrf) == "Group.2"] <- "Area"
names(average_pxrf)[names(average_pxrf) == "Group.3"] <- "Type"

# Change "Type" and "Area" to factors
average_pxrf$Type <- factor(average_pxrf$Type)
average_pxrf$Area <- factor(average_pxrf$Area)

# Removing columns that only have NA values
all_na <- function(x) any(!is.na(x))
pxrf_all <- average_pxrf %>% select_if(all_na)

# Removing columns if they contain any NA values at all
any_na <- function(x) all(!is.na(x))
pxrf_no_na <- average_pxrf %>% select_if(any_na)
```

**pxrf_no_na**: \
Good: Fe, Cu, Zn, Rb, Sr, Zr \
Neutral: Ti, CaO \
Inconclusive: Al2O2, SiO2, K2O,  Mn

```{r}
glimpse(pxrf_no_na)
```

**pxrf_all**: \
Good: Fe, Cu, Zn, Rb, Sr, Y, Zr, Nb \
Neutral: Ni, CaO, Ti, V \
Inconclusive: As, MgO, Al2O2, SiO2, S, Cl, K2O, Mn

```{r}
glimpse(pxrf_all)
```

### pXRF: correlations

Introduction

```{r}
# Drop the more dubious elements
pxrf_1 <- pxrf_no_na %>% select(-Al2O3)
pxrf_1 <- pxrf_1 %>% select(-CaO)

# Visualize correlations between elements
cor(pxrf_1[3:12]) %>% corrplot (type = "lower", tl.cex = 1, tl.pos="d", cl.pos="r")

# Optimal number of clusters in K-means analysis: 2
k_max <- 8
twcss <- sapply(1:k_max, function(k){kmeans(pxrf_1[3:12], k)$tot.withinss})
qplot(x = 1:k_max, y = twcss, geom = 'line')

# K-means clustering in ggpairs: 2 clusters
km <-kmeans(pxrf_1[3:12], centers = 2)
cluster_x <- as.factor(km$cluster)
ggpairs(pxrf_1, mapping = aes(col = cluster_x))
```

### pXRF: PCA

PCA with only the no-NA:s elements

```{r}
# Elements with no NA values at all
colnames(pxrf_no_na)

# PCA
pca_1 <- prcomp(pxrf_1[3:12])
summary(pca_1)

# PCA1 plots
biplot(pca_1, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"), main = "PCA Ashdod-Yam elements")

autoplot(pca_1, data=pxrf_1, colour='Area', shape = FALSE, label = TRUE,  main = "PCA Ashdod-Yam grouped by area")

autoplot(pca_1, data=pxrf_1, colour='Type', shape = FALSE, label = TRUE,  main = "PCA Ashdod-Yam grouped by sample type")
```
 
PCA with all the feasible elements

```{r}
# Elements that have at least one viable value
colnames(pxrf_all)

# PCA with NA:s converted to zeros
pxrf_all[is.na(pxrf_all)] <- 0

pca_2 <- prcomp(pxrf_all[3:22])
summary(pca_2)

# PCA2 plots
biplot(pca_2, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"), main = "PCA Ashdod-Yam more elements")

autoplot(pca_1, data=pxrf_1, colour='Area', shape = FALSE, label = TRUE,  main = "PCA Ashdod-Yam more elements grouped by area")

autoplot(pca_1, data=pxrf_1, colour='Type', shape = FALSE, label = TRUE,  main = "PCA Ashdod-Yam more elements grouped by sample type")

```

### pXRF: HCA

Introduction

```{r}
# HCA

# compute divisive hierarchical clustering
hc4 <- diana(pxrf_1[3:12])

# Divise coefficient; amount of clustering structure found
hc4$dc

# Diana dendrogram
pltree(hc4, cex = 1, hang = -1, main = "Dendrogram of diana")

# HCA dendrogam 1
dend <- 
    pxrf_1[3:12] %>%                    # data
    dist %>%                            # calculate a distance matrix
    hclust(method = "ward.D2") %>%      # hierarchical clustering 
    as.dendrogram %>%                   # turn the object into a dendrogram
    set("branches_k_color", k=5) %>% 
    set("branches_lwd", 1.2) %>%
    set("labels_col", c("red", "black", "blue"), k=5) %>%    
    set("labels_cex", 1) %>% 
    set("leaves_pch", NA) 
# plot the dend in usual "base" plotting engine:
plot(dend)


# HCA dendrogam 2: 
HCA.ward5 <- hclust(dist(pxrf_1[3:12], method="euclid"), method="ward.D2")
plot(HCA.ward5, main="Ward's method")
rect.hclust(HCA.ward5, k=5)
```

### Particle size

Introduction

```{r}
## Particle size
# Read and filter data
particle <- read.xlsx("data/grain_AY.xlsx", sep=";")

# Remove automatically created averages in order to include results from multiple runs for the same sample
particle <- particle %>% 
  filter(!grepl('Average', Sample)) %>% 
  filter(!grepl('test', Sample))

# Average by sample name
particle <- aggregate(particle, by=list(particle$Sample), FUN=mean)
particle <- particle %>% select(-Sample)
names(particle)[names(particle) == "Group.1"] <- "Sample"

# Saving the clean data as "AY_grain2"
write.xlsx(particle, file="data/grain/AY_grain2.xlsx")

# Manually added Type and Context columns to AY_grain2 data: "AY_grain3"
grain3 <- read.xlsx("data/grain/AY_grain3.xlsx", sep=";")

# Samples as rownames
rownames(grain3) <- grain3$Sample 

# Scaling clay-silt-sand portions to values between 0 and 1
## Otherwise the differences in low clay percentages are lost
normalize <- function(x, na.rm=TRUE){(x-min(x, na.rm=TRUE))/(max(x, na.rm=TRUE)-min(x, na.rm=TRUE))}

grain <- as.data.frame(apply(grain3[3:5], 2, normalize))

# Ternary plots

ggtern(data=grain, aes(x=Clay, y=Sand, z=Silt, color=grain3$Context)) +
  labs(title="Clay silt sand") +
  theme_rgbw() +
  theme_nomask() +
  geom_point(size=0) +
  geom_text(aes(label=rownames(grain)), size=3)

ggtern(data=grain, aes(x=Clay, y=Sand, z=Silt, color=grain3$Context)) +
  labs(title="Clay silt sand") +
  theme_rgbw() +
  theme_nomask() +
  geom_point(size=2)

ggtern(data=grain, aes(x=Clay, y=Sand, z=Silt, color=grain3$Type)) +
  labs(title="Clay silt sand") +
  theme_nomask() +
  geom_mask() +
  geom_point(size=2) 


```

### LOI

We did loss on ignition analysis both in the traditional manual method and with a modern TGA for comparison.

```{r}
loi <- read.xlsx("data/loi_AY.xlsx", sep=";")
tga <- read.xlsx("data/tga_AY.xlsx", sep=";")

glimpse(loi)
glimpse(tga)

# Remove rows containing words 'kekkila' (internal standard)
tga <- filter(tga, !grepl('kekkila', Name))

# Sample names as rownames
rownames(loi) <- loi$sample
rownames(tga) <- tga$Name

# Creating new columns for analysis by subtracting crucible mass;
# "dry" (sample mass after drying)
# "c550" (sample mass after 550 C)
# "c950" (sample mass after 950 C)
loi$dry <- (loi$dry_weight - loi$crucible)
loi$c550mass <- (loi$after_550_C - loi$crucible)
loi$c950mass <- (loi$after_950_C - loi$crucible)

# Converting values to percentages
loi$c550 <- (loi$dry - loi$c550mass)/(loi$dry) * 100
loi$c950 <- (loi$c550mass - loi$c950mass)/(loi$c550mass) * 100

# Total distribution of LOI in 550 C (1) and 950 C (2)
boxplot(loi$c550, loi$c950)

# boxplot.stats gives us the lower whisker or minimum, the interquartile range of the middle (the second and fourth values), the median (third value) and the upper whisker or maximum. 
boxplot.stats(loi$c950)

# The two high carbonate outliers are both AY-53 (two different measurements to make sure the first batch wasn't misleading)

# Removing duplicates from the next graph
rownames(loi)
loi <- subset(loi[1:47, ])

ggplot(loi, 
      aes(c550, c950, label = rownames(loi), colour = factor(context))) +
      geom_point(size=2, aes(shape = factor(type))) +
      geom_text_repel() + labs(title = "Organic vs. carbonate content") +
      xlab('Organic LOI (%)') +
      ylab('Carbonate LOI (%)') +
      theme(axis.title = element_text())
      




```

### Colorimetry

Introduction

```{r}

```



