---
author: "Uine Kailamäki"
output:
  html_document:
    theme: cosmo
    fig_caption: true
    fig_width: 12
    fig_height: 8 
  
---

From literature we can identify three rough classes of elements that could be used to analyze mudbricks from pXRF results: 

Good: Rb, Sr, Y, Zr, Nb, Th, Pb, Cu, Zn, Fe \
Neutral: V, Cr, Co, Ni, Ca, Ti \
Bad: Na, P, Ba (+ Ca, Al)

The "bad" elements were excluded from the pXRF results from the get-go. After scaling and averaging the remaining values we ended up with two analysis datasets: **"pxrf_all"** that contains all the elements that have non-zero results, and **"pxrf_no_na"**, in which all the elements that have even one NA result are omitted.  

#### Data preparation

```{r}

```

**pxrf_no_na**: \

```{r}
summary(pxrf_no_na)
```

**pxrf_all**: \

```{r}
summary(pxrf_all)
```

#### pXRF: Elemental correlations and K means cluster analysis

```{r}

```

#### pXRF: PCA

**PCA with only the no-NA:s elements**: \

```{r}

```

**PCA with all the feasible elements**: \

```{r}

```

#### pXRF: HCA

```{r}

```

