---
author: "Uine Kailamäki"
output:
  html_document:
    theme: cosmo
    fig_caption: true
    fig_width: 12
    fig_height: 8 
  
---

From literature we can identify three rough classes of elements that could be used to analyze mudbricks from pXRF results: 

Good: Rb, Sr, Y, Zr, Nb, Th, Pb, Cu, Zn, Fe \
Neutral: V, Cr, Co, Ni, Ca, Ti \
Bad: Na, P, Ba (+ Ca, Al)

The "bad" elements were excluded from the pXRF results from the get-go. After scaling and averaging the remaining values we ended up with two analysis datasets: **"pxrf_all"** that contains all the elements that have non-zero results, and **"pxrf_no_na"**, in which all the elements that have even one NA result are omitted.  

#### Data preparation

```{r}
# Libraries
library(openxlsx); 
library(dplyr);
library(ggfortify); # PCA autoplots
library(FactoMineR); # Fast PCA graphs
library(dendextend); # Dendrograms

# Read the raw data
pxrf <- read.xlsx("data/pxrf_israel.xlsx", sep=";")

# Remove any rows containing words 'ERROR' or 'TEST'
pxrf <- filter(pxrf, !grepl('TEST', Sample))
pxrf <- filter(pxrf, !grepl('ERROR', Sample))

# Remove unused columns
pxrf <- pxrf %>% 
        select(-Site) %>% 
        select(-'Other;name') %>% 
        select(-'File;#') %>% 
        select(-DateTime) %>% 
        select(-Application) %>% 
        select(-Method) %>% 
        select(-ElapsedTime) %>% 
        select(-'Cal;Check')
  
# Coercing "< LOD" -results to NA:s  
pxrf[4:93] <- pxrf[4:93] %>% mutate_if(is.character,as.numeric)

# Averaging results by "Sample", "Area" and "Type" columns
pxrf_averaged <- aggregate(pxrf, by = list(pxrf$Sample, pxrf$Area, pxrf$Type), FUN = mean)

# Removing old (now empty) columns
pxrf_averaged <- pxrf_averaged %>% select(-Sample)
pxrf_averaged <- pxrf_averaged %>% select(-Area)
pxrf_averaged <- pxrf_averaged %>% select(-Type)

# Assigning sample names as new row names 
rownames(pxrf_averaged) <- pxrf_averaged$Group.1
pxrf_averaged <- pxrf_averaged %>% select(-Group.1)

# Renaming the other two newly created groups back to "Type" and "Area"
names(pxrf_averaged)[names(pxrf_averaged) == "Group.2"] <- "Area"
names(pxrf_averaged)[names(pxrf_averaged) == "Group.3"] <- "Type"

# Removing error columns from the scaled analysis data set
pxrf_scaled <- pxrf_averaged %>% select(-ends_with("Err"))

# Scaling the data with standard z-score method
pxrf_scaled[3:47] <- scale(pxrf_scaled[3:47])

# Removing columns that only contain NA values
all_na <- function(x) any(!is.na(x))
pxrf_all <- pxrf_scaled %>% select_if(all_na)

```

**pxrf_averaged**: All averaged data including averaged errors (Uine's note: for later use)

```{r}
colnames(pxrf_averaged)
```

**pxrf_all**: All elements (the numbers below are the amount of NA values)

```{r}
colSums(is.na(pxrf_all))
```

**pxrf_final**: Final analysis data set with only the selected elements

```{r}
# Dropping the dubious elements and elements with too many NA values
pxrf_final <- pxrf_all %>% 
        select(-Al2O3) %>% 
        select(-CaO) %>% 
        select(-SiO2) %>%
        select(-Cl) %>%  
        select(-K2O) %>% 
        select(-Ni) %>% 
        select(-P2O5) %>%  
        select(-S) %>% 
        select(-Nb) 

# Transforming remaining NA values to zeros for analysis
pxrf_final[is.na(pxrf_final)] <- 0

# Final analysis data set        
colnames(pxrf_final)
```

**pxrf_MB**: Final analysis data set for  MUDBRICK samples only (= no floor or lime plaster)

```{r}
pxrf_MB <- pxrf_final[-c(1:7), ]
rownames(pxrf_MB)
```

#### pXRF: K means cluster analysis

```{r}
# PC autoplot with K-means clusters
set.seed(1)
autoplot(kmeans(pxrf_final[3:12], 7), data = pxrf_final, label = TRUE, label.size = 3)

```

#### pXRF: PCA

**PCA with final elements**: \

```{r}
# Elements
colnames(pxrf_final[3:12])

# PCA analysis
pca_1 <- prcomp(pxrf_final[3:12])
summary(pca_1)

# NO COPPER: Elements
pxrf_noCu <- pxrf_final[-7]
colnames(pxrf_noCu[3:11])

# NO COPPER: PCA analysis
pca_2 <- prcomp(pxrf_noCu[3:11])
summary(pca_2)

# PCA plots
biplot(pca_1, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"), main = "PCA Israel elements")
biplot(pca_2, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"), main = "PCA Israel elements (NO COPPER)")

autoplot(pca_1, data=pxrf_final, colour='Area', shape = FALSE, label = TRUE,  main = "PCA Israel grouped by area")
autoplot(pca_2, data=pxrf_noCu, colour='Area', shape = FALSE, label = TRUE,  main = "PCA Israel grouped by area (NO COPPER)")

autoplot(pca_1, data=pxrf_final, colour='Type', shape = FALSE, label = TRUE,  main = "PCA Israel grouped by sample type")
autoplot(pca_2, data=pxrf_noCu, colour='Type', shape = FALSE, label = TRUE,  main = "PCA Israel grouped by sample type (NO COPPER)")

PCA(pxrf_final[3:12])
PCA(pxrf_noCu[3:11])

```

**PCA with only mudbrick samples**: \

```{r}
# Elements
colnames(pxrf_MB[3:12])

# PCA analysis
pca_3 <- prcomp(pxrf_MB[3:12])
summary(pca_3)

# NO COPPER: Elements
pxrf_noCu_MB <- pxrf_MB[-7]
colnames(pxrf_noCu_MB[3:11])

# NO COPPER: PCA analysis
pca_4 <- prcomp(pxrf_noCu_MB[3:11])
summary(pca_4)

# PCA plots
biplot(pca_3, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"), main = "PCA Israel MB elements")
autoplot(pca_3, data=pxrf_MB, colour='Area', shape = FALSE, label = TRUE,  main = "PCA Israel MB grouped by area")
PCA(pxrf_MB[3:12])

biplot(pca_4, choices = 1:2, cex = c(1, 1.2), col = c("grey80", "deeppink2"), main = "PCA Israel MB elements")
autoplot(pca_4, data=pxrf_noCu_MB, colour='Area', shape = FALSE, label = TRUE,  main = "PCA Israel MB grouped by area")
PCA(pxrf_MB[3:12])

```

#### pXRF: HCA

```{r}
# HCA dendrogram, samples color coded by area:
dend <- 
    pxrf_MB %>%                      # data
    dist %>%                            # calculate a distance matrix
    hclust(method = "ward.D2") %>%      # hierarchical clustering 
    as.dendrogram %>%                   # turn the object into a dendrogram
    color_branches(k=4)

area <- as.factor(pxrf_MB[, 1])
n_area <- length(unique(area))
cols_a <- colorspace::rainbow_hcl(n_area, c = 70, l  = 50)
col_area <- cols_a[area]
labels_colors(dend) <- col_area[order.dendrogram(dend)]

plot(dend, main="HCA with sample areas")
rect.dendrogram(dend, 4, border = "Black")
legend("topright", legend = levels(area), fill = cols_a)

```


